---
title: "RR_Prog-1"
author: "lamcd"
date: "19/08/2020"
output: html_document
---
Reproducible Research - Programming Assignment 1
The .csv file is in the working directory

Initializing, reading and converting data - question 1

```{r}

knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library("knitr")
require("ggplot2")
library("plyr")

## read data and take a look at the file
activity <- read.csv("activity.csv")
summary(activity)

## convert Date column from char to type date variables
activity$date <- as.Date(activity$date)
str(activity)

```

Histo of total number of steps per day - question 2
Find how many missing values in the activity data
Remove missing values from activity column and create 'clean' per-day data
Aggregate the data (steps) by day and plot histogram - question 2

```{r}

## histo of total number of steps per day - question 2
activity_steps_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
hist(activity_steps_day$steps, xlab = "Steps per Day", main = "Total number of steps taken per day", col = "green")

## find how many missing values in the activity data
sum(is.na(activity$steps))

## remove missing values from activity column and create 'clean' per-day data
activity_rm <- activity[which(!is.na(activity$steps)),]

## aggregate the data (steps) by day and plot histogram - question 2
perday <- tapply(activity_rm$steps, activity_rm$date, sum)
hist(perday, 10, main = "Total number of steps taken per day", xlab = "steps per day", col = "green")

```

Calculate and display mean and median of steps-per-day - question 3 

```{r}

mean(perday)
median(perday)

## to analyse a daily mean activity we will aggregate the data (steps) by interval
## accross days in the data set
dailyact <- tapply(activity_rm$steps, activity_rm$interval, mean)

```

Plot the daily average number of steps - question 4

```{r}

## plot the daily average number of steps - question 4
plot(y = dailyact, x = names(dailyact), type = "l", xlab = "Intervals", 
     main = "Daily Activity Mean Pattern", ylab = "Average number of steps")

```

Find the 5-minute interval that contains (on average) the maximun number of steps
and that maximum value - question 5

```{r}

## find the 5-minute interval that contains (on average) the maximun number of steps
## and that maximum value - question 5
dailyact[dailyact==max(dailyact)]

## interval 835, with 206.2 steps

```
Define a strategy for imputing missing data - question 6 (start)
Determining where the missing values are in the data set

So, all the missing data is in the activity variable
Furthermore 2304 missing values is a considerable number - 
approximately 13%, so some fixing is needed.
An intuitive processe will be to replace the missing values with mean or median
for the day or the 5-minute interval

To start a new data set is created with missing values replaced by interval means
act_new <- activity
act_new[which(is.na(act_new$steps)),1]<-
  dailyact[as.character(act_new[which(is.na(act_new$steps)),3])]

Checking if there are remaining missing values
There are no remaining missing values

To decide about a strategy for imputing missing data 2 histos are plotted:
The original (removing NAs) and the new imputing interval mean - question 6
perday_new<-tapply(act_new$steps, act_new$date, sum)

```{r}

## define a strategy for imputing missing data - question 6 (start)
## determining where the missing values are in the data set
## missing data in the activity (steps) variable:
sum(is.na(activity$steps))
## missing data in the complete data set:
sum(is.na(activity))

## so, all the missing data is in the activity variable
## furthermore 2304 missing values is a considerable number - 
## approximately 13%, so some fixing is needed.
## an intuitive processe will be to replace the missing values with mean or median
## for the day or the 5-minute interval

## to start a new data set is created with missing values replaced by interval means
act_new <- activity
act_new[which(is.na(act_new$steps)),1]<-
  dailyact[as.character(act_new[which(is.na(act_new$steps)),3])]

## checking if there are remaining missing values
sum(is.na(act_new))
## there are no remaining missing values

## to decide about a strategy for imputing missing data 2 histos are plotted:
## the original (removing NAs) and the new imputing interval mean - question 6
perday_new<-tapply(act_new$steps, act_new$date, sum)

```
Histo's with same scale in Y-axis to facilitate comparison - question 7
Calculate and display mean and median after imputing NAs

So, imputing the missing data had little impact in either mean or median

```{r}

## histo's with same scale in Y-axis to facilitate comparison - question 7
par(mfrow=c(1,2))
hist(perday,10, main = "Total number of steps taken per day", xlab = "Steps"
     , ylim =c(0, 25))
abline(v = median(perday), col = 4, lwd = 4)
hist(perday_new,10, main = "Total number of steps taken per day  
     (missing values replaced with mean of interval)", xlab = "Steps",
     ylim =c(0, 25))
abline(v = median(perday_new), col = 4, lwd = 4)
dev.copy(png, file="plot5.png", height=480, width=480)

## calculate and display mean and median after imputing NAs
mean(perday_new)
median(perday_new)

## so, imputing the missing data had little impact in either mean or median

```

Studying the difference between activity during weekdays and weekends - question 8
to start create a factor with two levels - "weekend" and "weekday"
split the data into two sets (weekdays - weekends) to make separate plots
data is still steps over 5-minute interval
The classification (weekday/weekend) is adapted to the language setting 
(sunday = domingo / saturday = sabado)

Simple plotting with base system

```{r}

## studying the difference between activity during weekdays and weekends - question 8

## to start create a factor with two levels - "weekend" and "weekday"
## test adjusted to language: domingo / sabado
act_new$wd <- weekdays(act_new$date)
act_new$fwd <- as.factor(c("weekend", "weekday"))
act_new[act_new$wd == "domingo" | act_new$wd == "sabado" ,5] <- factor("weekend")
act_new[!(act_new$wd == "domingo" | act_new$wd == "sabado"),5 ] <- factor("weekday")

## split the data into two sets (weekdays - weekends) to make separate plots
## data is still steps over 5-minute interval
## simple plotting with base system
act_new_we <- subset(act_new, fwd == "weekend") 
act_new_wd <- subset(act_new, fwd == "weekday") 
dailyact_we <- tapply(act_new_we$steps, act_new_we$interval, mean)
dailyact_wd <- tapply(act_new_wd$steps, act_new_wd$interval, mean)

par(mfrow=c(2,1))
plot(y = dailyact_wd, x = names(dailyact_wd), type = "l", xlab = "5-Minute Interval", 
     main = "Daily Activity Pattern on Weekdays", ylab = "Average number of steps", 
     xlim = c(0, 2500), ylim = c(0, 250))
plot(y = dailyact_we, x = names(dailyact_we), type = "l", xlab = "5-Minute Interval", 
     main = "Daily Activity Pattern on Weekends", ylab = "Average number of steps", 
     xlim = c(0, 2500), ylim = c(0, 250))

## Analysis

```


So, the pattern of activity for weekdays is different from that of weenkends
On weekdays the activity seems to start earlier
On weekdays there seems to be a peak of activity in earlier hours then decreases
On weekends the activity spreads more uniformly across the day



